---
title: "Reproducible Research: Peer Assessment 1"
author: "Boris Blazej"
output: 
  html_document:
    keep_md: true
---

For the project we need some libraries 

```{r, libraries, results='hide'}
library(readr)
library(tidyverse)
```


## Loading and preprocessing the data

First the source data is unzipped to the data directory

```{r unzip}
zipFile <- "./activity.zip"
dataPath <- "./data"
if (!file.exists(dataPath)) {
    unzip(zipFile, exdir = dataPath)
}

```
We read the data and parse the date format: 

```{r read_csv}
activity <- read_csv("data/activity.csv", 
    col_types = cols(date = col_date(format = "%Y-%m-%d"), 
        steps = col_double()))
```




## What is mean total number of steps taken per day?

We calculate the total number of steps by day and draw a histogram: 

```{r histogram1}
daily <- activity %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm = TRUE))
hist(daily$total_steps, breaks = 10,
     main = "Total Number of Steps per Day")
```

```{r mean_median1}
mtns <- mean(daily$total_steps)
mtns
medtns <- median(daily$total_steps)
medtns
```

The mean total number of steps taken per day is **`r as.integer(mtns)`**, the median is at **`r as.integer(medtns)`**.




## What is the average daily activity pattern?

Averaging steps per intervall over all available days shows us the average activity pattern:

```{r pattern}
pattern <- activity %>%
    group_by(interval) %>%
    summarise(avg_steps = mean(steps, na.rm = TRUE))
with(pattern, plot(x=interval, y = avg_steps, type = "l"))

```

```{r most_act}
most_act <- pattern[which.max(pattern$avg_steps),]
most_act
```

The most active interval is the one at **`r most_act$interval`** seconds with ~**`r as.integer(most_act$avg_steps)` steps**.


## Imputing missing values

To fill NA gaps we will replace them by the mean of the respective interval, which we have already available as the "overall activity pattern" (variable: pattern).


```{r replace_NA}
activity1 <- activity

for(r in 1:nrow(activity1)){
    if(is.na(activity1$steps[r])) {
        activity1$steps[r] <- 
            pattern[which(pattern$interval == activity1$interval[r]),]$avg_steps
    }
}

sum(is.na(activity1$steps))
    
```

The new data set doe not contain NA's anymore. So we draw again a histogram ...

```{r histogram2}
daily1 <- activity1 %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm = TRUE))
hist(daily1$total_steps, breaks = 10,
     main = "Total Number of Steps per Day")
```

... and calculate mean and median:


```{r mean_median2}
mtns1 <- mean(daily1$total_steps)
mtns1
medtns1 <- median(daily1$total_steps)
medtns1
```

The mean total number of steps taken per day is **`r as.integer(mtns1)`**, the median is at **`r as.integer(medtns1)`**.




## Are there differences in activity patterns between weekdays and weekends?

First we caclulate the day and whether it is a weekday or weekend and reproduce the activity pattern for each:

```{r weekdays}
activity1 <- activity1 %>%
    mutate(day = weekdays(date), 
           daytype = ifelse(day %in% c("Montag", "Dienstag",
                                       "Mittwoch", "Donnerstag",
                                       "Freitag"), 
                            "weekday", "weekend"))

pattern1 <- activity1 %>%
    group_by(interval, daytype) %>%
    summarise(avg_steps = mean(steps, na.rm = TRUE))

```

Finally, we plot the different patterns:

```{r weekday_pattern}
ggplot(data = pattern1, aes(interval, avg_steps)) +
    geom_line()  +
    facet_grid(daytype ~ .)
```


